---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

</br>
Libraries:

```{r, echo=TRUE, message=FALSE}
library(dplyr)
library(ggplot2)
```

</br>
Code to load data, with `readr::col_types` to parse columns' values in the correct format.

```{r, echo=TRUE}
raw_data <- 
    readr::read_delim(
    file = "activity.zip",
    delim = ",",
    na = "NA",
    col_names = TRUE,
    col_types    = readr::cols(
        steps    = readr::col_double(),
        date     = readr::col_date(format = ""),
        interval = readr::col_double()
    )
  )
```

</br>

## What is mean total number of steps taken per day?

</br>

#### 1. Make a histogram of the total number of steps taken each day

Used `dplyr::filter()` to filter out NAs, `dplyr::group_by()` and `dplyr::summarise()` to respectively group by date (day) and calculate the sum of steps for each group. Then `ggplot2::geom_histogram()` to plot results.

```{r, echo=TRUE, message=FALSE, fig.align='center'}
# prepare and compute data

daily_tot_steps <- 
    raw_data %>%
    filter(!is.na(steps)) %>%
    group_by(date) %>%
    summarise(tot_steps = sum(steps))



# plot the results

ggplot(daily_tot_steps) + 
geom_histogram(mapping = aes(x = tot_steps), binwidth = 750, fill = "gray66", color = "white") +
theme_bw() +
labs(x = "", y = "Count", title = "Histogram of the total number of steps taken each day")

```

</br>

#### 2. Calculate and report the **mean** and **median** total number of steps taken per day

</br>

```{r, echo=TRUE}
# penalize scientific notations
options(scipen = 9999)

# calculate mean and median

daily_steps_mean <- with(daily_tot_steps, round(mean(tot_steps, na.rm = TRUE), digits = 2))
daily_steps_med  <- median(daily_tot_steps$tot_steps, na.rm = TRUE)
```

```{r, echo=TRUE}
# mean
daily_steps_mean
```

```{r, echo=TRUE}
# median
daily_steps_med
```
</br>

The mean is **`r daily_steps_mean`** and the median is **`r daily_steps_med`**.

</br>

## What is the average daily activity pattern?

</br>

#### 1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

Used `dplyr::filter()` to remove NAs, `dplyr::group_by()` and `dplyr::summarise()` to respectively group by interval and calculate the average of steps for each group. Then `ggplot2::geom_line()` to plot results.

```{r, echo=TRUE, message=FALSE}
int_avg_steps <-
    raw_data %>%
    filter(!is.na(steps)) %>%
    group_by(interval) %>%
    summarise(avg_steps = mean(steps))


ggplot(int_avg_steps) +
geom_line(mapping = aes(x = interval, y = avg_steps))
```

</br>

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

</br>

```{r, echo=TRUE, message=FALSE}
max_int <- as.numeric(int_avg_steps[which.max(int_avg_steps$avg_steps), 1])

# maximum 5-min interval
max_int
```

</br>

## Imputing missing values

</br>

#### 1. Calculate and report the total number of missing values in the dataset

</br>

```{r, echo=TRUE}
miss_data <- sum(is.na(raw_data$steps))

# number of missing values
miss_data
```

</br>

#### 2. Devise a strategy for filling in all of the missing values in the dataset.

I replaced all missing values (NAs) with the correspondent 5-minute interval average, already calculated in the previous question.

</br>

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

</br>

```{r, echo=TRUE}
fill_data <- raw_data

fill_data[is.na(fill_data$steps), 1] <- 
    int_avg_steps[int_avg_steps$interval %in% fill_data$interval, 2]
```

</br>

#### 4. Make a histogram of the total number of steps taken each day

</br>

```{r, echo=TRUE, fig.align='center'}
# prepare and compute data

fill_daily_steps <- 
  fill_data %>%
  group_by(date) %>%
  summarise(tot_steps = sum(steps, na.rm = TRUE))


# plot results

ggplot(fill_daily_steps) + 
  geom_histogram(mapping = aes(x = tot_steps), binwidth = 750, fill = "gray66", color = "white") +
  theme_bw() +
  labs(x = "", y = "Count", title = "Histogram of the total number of steps taken each day (filled)")
```

</br>

####  Calculate and report the mean and median total number of steps taken per day.

</br>

```{r}
# calculate meand and median

fill_steps_mean <- with(fill_daily_steps, round(mean(tot_steps, na.rm = TRUE), digits = 2))
fill_steps_med  <- median(fill_daily_steps$tot_steps, na.rm = TRUE)
```
```{r}
fill_steps_mean
```
```{r}
fill_steps_med
```



## Are there differences in activity patterns between weekdays and weekends?


